---
output: github_document
---

## Bayesian Quantile Methylation Analysis

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```



## Overview

BQMA (Bayesian Quantile Methylation Analysis) implements Bayesian sparse regression for Illumina methylation arrays. It provides posterior inclusion probabilities (PIP) for each CpG probe, quantile-specific effect profiles for continuous phenotypes, and a probit model for binary phenotypes. The spike-slab prior performs variable selection directly, returning a probability of association rather than a p-value.
This vignette demonstrates the full workflow on simulated data that mimics a real methylation array experiment, then shows how to compare results with limma.

## Installation

```{r install, eval = FALSE}
#if (!requireNamespace("BiocManager", quietly = TRUE))
#  install.packages("BiocManager")
#BiocManager::install("BQMA")
```

## Simulating methylation data

We simulate a dataset with 200 samples and 100 CpG probes. Ten probes are truly associated with a binary phenotype (case/control). The remaining 90 probes are null. This setup mimics a small methylation array experiment and allows us to evaluate method performance against known ground truth.

```{r simulate}
library(BQMA)

set.seed(42)
n <- 200
p <- 100

X <- matrix(rnorm(n * p), nrow = n, ncol = p)
colnames(X) <- paste0("cg", seq_len(p))

# Ten truly associated probes with moderate effects
beta_true            <- rep(0, p)
beta_true[1:5]       <- c(0.8, -0.8, 0.6, -0.6, 0.7)
beta_true[6:10]      <- c(-0.7, 0.5, -0.5, 0.6, -0.6)

# Binary phenotype generated from a probit model
eta <- X %*% beta_true
y   <- rbinom(n, 1, pnorm(eta))

cat("Cases    :", sum(y == 1), "\n")
cat("Controls :", sum(y == 0), "\n")
cat("Signal probes  :", sum(beta_true != 0), "\n")
cat("Null probes    :", sum(beta_true == 0), "\n")
```

The correlation between the true signal probes and the phenotype confirms that the simulated signal is detectable but moderate, as expected in real EWAS data.

```{r check_signal}
cor_true  <- mean(abs(cor(X[, 1:10],  y)))
cor_false <- mean(abs(cor(X[, 11:100], y)))

cat("Mean correlation - signal probes:", round(cor_true,  3), "\n")
cat("Mean correlation - null probes  :", round(cor_false, 3), "\n")
```

## Fitting BQMA for a binary phenotype

The main function is `BQMA()`. For a binary phenotype coded as 0/1 we use `family = "binary"`, which fits a Bayesian probit model with a spike-slab prior using Albert and Chib (1993) data augmentation and the George and McCulloch (1993) marginal Bayes factor for variable selection.

The `expected_incl` argument sets the prior probability that any given probe is associated with the phenotype. The default of 0.05 encodes the expectation that roughly 5 percent of tested probes are truly associated, which is a reasonable assumption for most EWAS.

```{r fit_bqma, cache = TRUE}
fit <- BQMA(
  X             = X,
  y             = y,
  family        = "binary",
  n_iter        = 3000L,
  burnin        = 1000L,
  expected_incl = 0.05,
  verbose       = FALSE
)

show(fit)
```

## Extracting results

The main output is the posterior inclusion probability (PIP) for each probe. A PIP close to 1 means the data strongly support inclusion of that probe in the model. A PIP close to 0 means the probe is likely null. We use a threshold of 0.5 as a default decision boundary.

```{r extract, warning=F}
library(minfi)
pip  <- getPIP(fit)[, 1]
beta <- BQMA::getBeta(fit)[, 1]
ci   <- getCI(fit)

cat("Probes with PIP > 0.5:", sum(pip > 0.5), "\n")
cat("Probes with PIP > 0.8:", sum(pip > 0.8), "\n")

# Top 10 probes ordered by PIP
top10 <- order(pip, decreasing = TRUE)[1:10]
# estrai CI correttamente dall'array 3D
ci_mat <- getCI(fit)[, 1, ]   # CpGs x 2 -- prendi il primo tau

results <- data.frame(
  probe   = names(pip)[top10],
  pip     = round(pip[top10], 3),
  beta    = round(beta[top10], 3),
  ci_low  = round(ci_mat[top10, 1], 3),
  ci_high = round(ci_mat[top10, 2], 3),
  is_true = names(pip)[top10] %in% paste0("cg", 1:10)
)

print(results)

```
