---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# BQMA: Bayesian Quantile Methylation Analysis

BQMA implements Bayesian sparse regression for Illumina methylation
arrays (450k / EPIC). It returns posterior inclusion probabilities
(PIP) per CpG probe, with a probit spike-slab model for binary
phenotypes (Albert & Chib 1993; George & McCulloch 1993) and an
Asymmetric Laplace quantile regression model for continuous
phenotypes (Kozumi & Kobayashi 2011). It also annotates probes with
QC flags (SNP at CpG, cross-reactive, sex chromosome) drawn from the
Illumina manifest and the Chen / McCartney probe lists.

## Installation

```{r install, eval = FALSE}
# if (!requireNamespace("BiocManager", quietly = TRUE))
#   install.packages("BiocManager")
# BiocManager::install("BQMA")
```


## Quick start


A minimal end-to-end example: simulate a binary EWAS, fit BQMA, extract
PIPs, then annotate the result.

```{r quickstart, message = FALSE, warning = FALSE}
library(BQMA)
library(IlluminaHumanMethylation450kanno.ilmn12.hg19)

set.seed(42)
n <- 200
p <- 100

X <- matrix(rnorm(n * p), nrow = n, ncol = p)
colnames(X) <- paste0("cg", seq_len(p))

# 10 truly associated probes
beta_true       <- rep(0, p)
beta_true[1:5]  <- c( 0.8, -0.8,  0.6, -0.6,  0.7)
beta_true[6:10] <- c(-0.7,  0.5, -0.5,  0.6, -0.6)

# Binary phenotype from a probit model
y <- rbinom(n, 1, pnorm(X %*% beta_true))

fit <- BQMA(X, y,
            family        = "binary",
            n_iter        = 3000L,
            burnin        = 1000L,
            expected_incl = 0.05,
            verbose       = FALSE)

show(fit)
```

## Extracting results

```{r extract}
pip  <- getPIP(fit)[, 1]
beta <- BQMA::getBeta(fit)[, 1]   # qualified to avoid conflict with minfi::getBeta
ci   <- getCI(fit)[, 1, ]         # CpGs x 2, first tau

cat("Probes with PIP > 0.5:", sum(pip > 0.5), "\n")
cat("Probes with PIP > 0.8:", sum(pip > 0.8), "\n")

top10 <- order(pip, decreasing = TRUE)[1:10]

data.frame(
  probe   = names(pip)[top10],
  pip     = round(pip[top10],     3),
  beta    = round(beta[top10],    3),
  ci_low  = round(ci[top10, 1],   3),
  ci_high = round(ci[top10, 2],   3),
  is_true = names(pip)[top10] %in% paste0("cg", 1:10)
)
```

## Annotating probes with QC flags

`annotate_results()` returns a tidy `data.frame` with the posterior
summaries plus per-probe QC classification (`CLEAN_CpG`,
`FLAGGED_snp_body`, `PROBLEMATIC`) and genomic context (gene, chr, pos).
Use real Illumina probe IDs (cgXXXXXXXX from the 450k or EPIC manifest)
for the annotation lookup to return non-NA flags.

```{r}
ann_df <- annotate_results(fit, array = "450k")
table(ann_df$probe_class)


# Filter to reliable hits
reliable <- ann_df[ann_df$pip >= 0.5 &
                   ann_df$probe_class == "CLEAN_CpG", ]
```

To attach the classification onto the `BQMAResult` object (so that
`show(fit)` prints the QC summary), use `add_annotation()`:

```{r}
fit <- add_annotation(fit, array = "450k")
show(fit)
```

## Documentation

See the package vignette for the full workflow on real probe IDs from
`minfiData`, plus a comparison against limma:

```{r}
#vignette("How-to-use-BMQA_new", package = "BQMA")
```

## References

- Albert JH, Chib S (1993). Bayesian analysis of binary and
  polychotomous response data. *JASA* 88(422), 669–679.
- George EI, McCulloch RE (1993). Variable selection via Gibbs
  sampling. *JASA* 88(423), 881–889.
- Kozumi H, Kobayashi G (2011). Gibbs sampling methods for Bayesian
  quantile regression. *J. Stat. Comput. Simul.* 81(11), 1565–1578.
- Chen YA et al. (2013). Discovery of cross-reactive probes and
  polymorphic CpGs in the Illumina HumanMethylation450 microarray.
  *Epigenetics* 8(2), 203–209.
- McCartney DL et al. (2016). Identification of polymorphic and
  off-target probe binding sites on the Illumina MethylationEPIC
  BeadChip. *Genomics Data* 9, 22–24.
