---
title: "Vignette for BMQA"
author: "Giulia Nicole Baldrighi"
date: "2026-04-16"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 5
)
```

## Overview

BQMA (Bayesian Quantile Methylation Analysis) implements Bayesian sparse
regression for Illumina methylation arrays. It provides posterior inclusion
probabilities (PIP) for each CpG probe, quantile-specific effect profiles
for continuous phenotypes, and a probit model for binary phenotypes.
The spike-slab prior performs variable selection directly, returning a
probability of association rather than a p-value.

This vignette demonstrates the full workflow on simulated data that mimics
a real methylation array experiment, then shows how to compare results
with limma.

## Installation

```{r install, eval = FALSE}
#if (!requireNamespace("BiocManager", quietly = TRUE))
#  install.packages("BiocManager")
#BiocManager::install("BQMA")
```

## Simulating methylation data

We simulate a dataset with 200 samples and 100 CpG probes. Ten probes
are truly associated with a binary phenotype (case/control). The remaining
90 probes are null. This setup mimics a small methylation array experiment
and allows us to evaluate method performance against known ground truth.

```{r simulate}
library(BQMA)

set.seed(42)
n <- 200
p <- 100

X <- matrix(rnorm(n * p), nrow = n, ncol = p)
colnames(X) <- paste0("cg", seq_len(p))

# Ten truly associated probes with moderate effects
beta_true            <- rep(0, p)
beta_true[1:5]       <- c(0.8, -0.8, 0.6, -0.6, 0.7)
beta_true[6:10]      <- c(-0.7, 0.5, -0.5, 0.6, -0.6)

# Binary phenotype generated from a probit model
eta <- X %*% beta_true
y   <- rbinom(n, 1, pnorm(eta))

cat("Cases    :", sum(y == 1), "\n")
cat("Controls :", sum(y == 0), "\n")
cat("Signal probes  :", sum(beta_true != 0), "\n")
cat("Null probes    :", sum(beta_true == 0), "\n")
```

The correlation between the true signal probes and the phenotype confirms
that the simulated signal is detectable but moderate, as expected in
real EWAS data.

```{r check_signal}
cor_true  <- mean(abs(cor(X[, 1:10],  y)))
cor_false <- mean(abs(cor(X[, 11:100], y)))

cat("Mean correlation - signal probes:", round(cor_true,  3), "\n")
cat("Mean correlation - null probes  :", round(cor_false, 3), "\n")
```

## Fitting BQMA for a binary phenotype

The main function is `BQMA()`. For a binary phenotype coded as 0/1 we
use `family = "binary"`, which fits a Bayesian probit model with a
spike-slab prior using Albert and Chib (1993) data augmentation and
the George and McCulloch (1993) marginal Bayes factor for variable
selection.

The `expected_incl` argument sets the prior probability that any given
probe is associated with the phenotype. The default of 0.05 encodes
the expectation that roughly 5 percent of tested probes are truly
associated, which is a reasonable assumption for most EWAS.

```{r fit_bqma, cache = TRUE}
fit <- BQMA(
  X             = X,
  y             = y,
  family        = "binary",
  n_iter        = 3000L,
  burnin        = 1000L,
  expected_incl = 0.05,
  verbose       = FALSE
)

show(fit)
```

## Extracting results

The main output is the posterior inclusion probability (PIP) for each
probe. A PIP close to 1 means the data strongly support inclusion of
that probe in the model. A PIP close to 0 means the probe is likely null.
We use a threshold of 0.5 as a default decision boundary.

```{r extract}
library(minfi)
pip  <- getPIP(fit)[, 1]
beta <- BQMA::getBeta(fit)[, 1]
ci   <- getCI(fit)

cat("Probes with PIP > 0.5:", sum(pip > 0.5), "\n")
cat("Probes with PIP > 0.8:", sum(pip > 0.8), "\n")

# Top 10 probes ordered by PIP
top10 <- order(pip, decreasing = TRUE)[1:10]
# estrai CI correttamente dall'array 3D
ci_mat <- getCI(fit)[, 1, ]   # CpGs x 2 -- prendi il primo tau

results <- data.frame(
  probe   = names(pip)[top10],
  pip     = round(pip[top10], 3),
  beta    = round(beta[top10], 3),
  ci_low  = round(ci_mat[top10, 1], 3),
  ci_high = round(ci_mat[top10, 2], 3),
  is_true = names(pip)[top10] %in% paste0("cg", 1:10)
)

print(results)

```

## PIP distribution

A well-calibrated BQMA fit produces a bimodal PIP distribution: most
probes cluster near zero (null probes) and a small number cluster near
one (signal probes). This separation is the key advantage over frequentist
p-values, which do not provide a direct probability of association.

```{r pip_hist, fig.cap = "PIP distribution. Purple bars indicate probes with PIP above 0.5."}
hist(pip,
     breaks = 30,
     col    = ifelse(
       seq(min(pip), max(pip), length.out = 30) > 0.5,
       "purple", "grey80"
     ),
     main = "Posterior inclusion probability distribution",
     xlab = "PIP")
abline(v = 0.5, lty = 2, col = "red", lwd = 2)
```

## Comparison with limma

We compare BQMA against limma, the standard frequentist tool for
differential methylation analysis. On simulated data with known ground
truth we can directly measure how many of the 10 truly associated probes
each method recovers in its top 10 results.

```{r limma_comparison}
if (!requireNamespace("limma", quietly = TRUE))
  BiocManager::install("limma")
library(limma)

design    <- model.matrix(~ y)
fit_limma <- lmFit(t(X), design)
fit_limma <- eBayes(fit_limma)
limma_res <- topTable(fit_limma, coef = 2,
                       number = Inf, sort.by = "P")

bqma_top10  <- names(pip)[order(pip, decreasing = TRUE)][1:10]
limma_top10 <- rownames(limma_res)[1:10]

cat("BQMA  - true probes in top 10:",
    sum(bqma_top10  %in% paste0("cg", 1:10)), "/10\n")
cat("limma - true probes in top 10:",
    sum(limma_top10 %in% paste0("cg", 1:10)), "/10\n")
cat("Overlap between methods      :",
    length(intersect(bqma_top10, limma_top10)), "\n")

comparison <- data.frame(
  rank       = 1:10,
  BQMA       = bqma_top10,
  BQMA_true  = bqma_top10  %in% paste0("cg", 1:10),
  limma      = limma_top10,
  limma_true = limma_top10 %in% paste0("cg", 1:10),
  in_common  = bqma_top10 %in% limma_top10
)

print(comparison)
```

## PIP versus limma p-value

The scatter plot below illustrates the relationship between BQMA PIP
and limma p-value. Probes with high PIP tend to have low p-values, but
the two measures are not equivalent. BQMA can assign high PIP to probes
that limma ranks poorly when their signal is consistent across the
posterior but does not reach the limma significance threshold after
multiple testing correction. These are the unique discoveries enabled
by the Bayesian framework.

```{r pip_vs_pvalue, fig.cap = "BQMA PIP versus limma p-value. Purple filled points are truly associated probes. The red dashed line marks PIP = 0.5. The blue dashed line marks nominal p = 0.05."}
plot_df <- data.frame(
  pip      = pip,
  neg_logp = -log10(limma_res[names(pip), "P.Value"]),
  is_true  = names(pip) %in% paste0("cg", 1:10)
)

plot(plot_df$neg_logp, plot_df$pip,
     col  = ifelse(plot_df$is_true, "purple", "grey60"),
     pch  = ifelse(plot_df$is_true, 19, 1),
     xlab = "-log10(p-value limma)",
     ylab = "PIP (BQMA)",
     main = "BQMA PIP versus limma p-value",
     ylim = c(0, 1))
abline(h = 0.5, lty = 2, col = "red",  lwd = 2)
abline(v = -log10(0.05), lty = 2, col = "blue", lwd = 2)
legend("topleft",
       legend = c("Signal probe", "Null probe",
                  "PIP = 0.5", "p = 0.05"),
       col    = c("purple", "grey60", "red", "blue"),
       pch    = c(19, 1, NA, NA),
       lty    = c(NA, NA, 2, 2),
       bty    = "n")
```

## Interpreting the results

The key outputs for downstream analysis are:

- `pip` values above 0.5 indicate probable association. Use 0.8 for
  a more stringent threshold analogous to genome-wide significance.
- `beta` values give the direction and magnitude of the association
  on the standardised M-value scale.
- `ci` values give the 95 percent credible interval for each probe.
  A credible interval that excludes zero is consistent with a
  non-null effect but PIP is the primary evidence measure.

```{r final_summary}
sig_probes <- names(pip)[pip >= 0.5]

cat("Significant probes (PIP >= 0.5):", length(sig_probes), "\n\n")

ci_mat <- getCI(fit)[, 1, ]

summary_df <- data.frame(
  probe     = sig_probes,
  pip       = round(pip[sig_probes], 3),
  beta      = round(beta[sig_probes], 3),
  ci_low    = round(ci_mat[sig_probes, 1], 3),
  ci_high   = round(ci_mat[sig_probes, 2], 3),
  direction = ifelse(beta[sig_probes] > 0,
                     "hypermethylated", "hypomethylated")
)

summary_df <- summary_df[order(summary_df$pip, decreasing = TRUE), ]
print(summary_df)
```

