---
title: "Vignette for BQMA"
author: "Giulia Nicole Baldrighi"
date: "2026-04-16"
output: html_document
vignette: >
  %\VignetteIndexEntry{Vignette for BQMA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 5
)
```

## Overview

BQMA (Bayesian Quantile Methylation Analysis) implements Bayesian sparse
regression for Illumina methylation arrays. It provides posterior
inclusion probabilities (PIP) for each CpG probe, quantile-specific
effect profiles for continuous phenotypes, and a probit model for binary
phenotypes. The spike-slab prior performs variable selection directly,
returning a probability of association rather than a p-value.

This vignette demonstrates the full workflow using **real Illumina 450k
probe identifiers** drawn from the `minfiData` example dataset, with a
controlled simulated design matrix and phenotype on top of those
identifiers. Real probe IDs are needed so that the annotation step
(SNP flags, cross-reactive filtering, chromosome lookup) actually
returns information; the simulation on `X` and `y` provides a known
ground truth so that BQMA's recovery rate can be measured against a
benchmark.

## Installation

```{r install, eval = FALSE}
# if (!requireNamespace("BiocManager", quietly = TRUE))
#   install.packages("BiocManager")
# BiocManager::install("BQMA")
```

## Loading real probe identifiers

We load the `MsetEx` dataset from `minfiData`, a 6-sample 450k example
shipped with Bioconductor, purely to obtain a realistic set of probe
IDs and to subset to the most variable probes. The actual model fit
will use a simulated design matrix on top of these identifiers; we are
*not* analysing the `MsetEx` samples themselves, since 6 samples are
too few for a meaningful EWAS.

```{r load_probes, message = FALSE}
library(BQMA)
library(minfi)
library(minfiData)
library(IlluminaHumanMethylation450kanno.ilmn12.hg19)

# Compute per-probe variance across the 6 MsetEx samples and pick the
# top 200 most variable probes. These are real cgXXXXXXXX IDs from the
# 450k manifest, including the realistic mix of clean / SNP-affected /
# cross-reactive / sex-chromosome probes that any real EWAS encounters.
M_full   <- minfi::getM(MsetEx)
probe_var <- matrixStats::rowVars(M_full, useNames = TRUE)
top_probes <- names(sort(probe_var, decreasing = TRUE))[seq_len(200)]

# Drop probes with NA variance (rare, e.g. all-zero rows)
top_probes <- top_probes[!is.na(top_probes)]
length(top_probes)
head(top_probes)
```

## Simulating phenotype and signal on real probes

We simulate a dataset of 200 samples on the 200 selected real probes.
Ten probes (chosen at random from the pool) are truly associated with a
binary phenotype; the remaining 190 are null. The phenotype is
generated from a probit model.

This gives us:

- a controlled ground truth (we know which probes are signal),
- real probe identifiers (so `annotate_results()` returns real flags),
- enough samples for the Bayesian model to converge (n = 200, p = 200).

```{r simulate}
set.seed(42)
n <- 200
p <- length(top_probes)

X <- matrix(rnorm(n * p), nrow = n, ncol = p)
colnames(X) <- top_probes

# Pick 10 random probes as the true signal
true_idx    <- sample(seq_len(p), 10)
true_probes <- colnames(X)[true_idx]

beta_true <- rep(0, p)
beta_true[true_idx] <- c( 0.8, -0.8,  0.6, -0.6,  0.7,
                         -0.7,  0.5, -0.5,  0.6, -0.6)

# Binary phenotype generated from a probit model.
eta <- X %*% beta_true
y   <- rbinom(n, 1, pnorm(eta))

cat("Cases    :", sum(y == 1), "\n")
cat("Controls :", sum(y == 0), "\n")
cat("Signal probes :", length(true_probes), "\n")
cat("Null probes   :", p - length(true_probes), "\n")
```

The correlation between true signal probes and the phenotype confirms
that the simulated signal is detectable but moderate, as expected in
real EWAS data.

```{r check_signal}
null_idx <- setdiff(seq_len(p), true_idx)
cor_true  <- mean(abs(cor(X[, true_idx], y)))
cor_false <- mean(abs(cor(X[, null_idx], y)))

cat("Mean correlation - signal probes:", round(cor_true,  3), "\n")
cat("Mean correlation - null probes  :", round(cor_false, 3), "\n")
```

## Fitting BQMA for a binary phenotype

The main function is `BQMA()`. For a binary phenotype coded as 0/1 we
use `family = "binary"`, which fits a Bayesian probit model with a
spike-slab prior using Albert and Chib (1993) data augmentation and the
George and McCulloch (1993) marginal Bayes factor for variable
selection.

The `expected_incl` argument sets the prior probability that any given
probe is associated with the phenotype. The default of 0.05 encodes the
expectation that roughly 5 % of tested probes are truly associated,
which is a reasonable assumption for most EWAS.

```{r fit_bqma, cache = TRUE}
fit <- BQMA(
  X             = X,
  y             = y,
  family        = "binary",
  n_iter        = 3000L,
  burnin        = 1000L,
  expected_incl = 0.05,
  verbose       = FALSE
)

show(fit)
```

## Extracting results

The main output is the posterior inclusion probability (PIP) for each
probe. A PIP close to 1 means the data strongly support inclusion of
that probe; a PIP close to 0 means the probe is likely null. We use
0.5 as a default decision threshold.

```{r extract}
pip  <- getPIP(fit)[, 1]
beta <- getBeta(fit)[, 1]
ci   <- getCI(fit)[, 1, ]   # CpGs x 2, first tau

cat("Probes with PIP > 0.5:", sum(pip > 0.5), "\n")
cat("Probes with PIP > 0.8:", sum(pip > 0.8), "\n")

# Top 10 probes ordered by PIP
top10 <- order(pip, decreasing = TRUE)[1:10]

results <- data.frame(
  probe   = names(pip)[top10],
  pip     = round(pip[top10],   3),
  beta    = round(beta[top10],  3),
  ci_low  = round(ci[top10, 1], 3),
  ci_high = round(ci[top10, 2], 3),
  is_true = names(pip)[top10] %in% true_probes
)
print(results)
```

## PIP distribution

A well-calibrated BQMA fit produces a bimodal PIP distribution: most
probes cluster near zero (null) and a small number cluster near one
(signal). This separation is the key advantage over frequentist
p-values, which do not provide a direct probability of association.

```{r pip_hist, fig.cap = "PIP distribution. Purple bars indicate probes with PIP above 0.5."}
hist(pip,
     breaks = 30,
     col    = ifelse(
       seq(min(pip), max(pip), length.out = 30) > 0.5,
       "purple", "grey80"
     ),
     main = "Posterior inclusion probability distribution",
     xlab = "PIP")
abline(v = 0.5, lty = 2, col = "red", lwd = 2)
```

## Comparison with limma

We compare BQMA against limma, the standard frequentist tool for
differential methylation analysis. With known ground truth we can
measure how many of the 10 truly associated probes each method recovers
in its top 10 results.

```{r limma_comparison, message = FALSE}
if (!requireNamespace("limma", quietly = TRUE))
  stop("Install 'limma' to run this comparison: BiocManager::install('limma')")
library(limma)

design    <- model.matrix(~ y)
fit_limma <- lmFit(t(X), design)
fit_limma <- eBayes(fit_limma)
limma_res <- topTable(fit_limma, coef = 2,
                      number = Inf, sort.by = "P")

bqma_top10  <- names(pip)[order(pip, decreasing = TRUE)][1:10]
limma_top10 <- rownames(limma_res)[1:10]

cat("BQMA  - true probes in top 10:",
    sum(bqma_top10  %in% true_probes), "/10\n")
cat("limma - true probes in top 10:",
    sum(limma_top10 %in% true_probes), "/10\n")
cat("Overlap between methods      :",
    length(intersect(bqma_top10, limma_top10)), "\n")

comparison <- data.frame(
  rank       = seq_len(10),
  BQMA       = bqma_top10,
  BQMA_true  = bqma_top10  %in% true_probes,
  limma      = limma_top10,
  limma_true = limma_top10 %in% true_probes,
  in_common  = bqma_top10 %in% limma_top10
)
print(comparison)
```

## PIP versus limma p-value

The scatter plot below illustrates the relationship between BQMA PIP
and limma p-value. Probes with high PIP tend to have low p-values, but
the two measures are not equivalent. BQMA can assign high PIP to probes
that limma ranks poorly when their signal is consistent across the
posterior but does not reach the limma significance threshold after
multiple testing correction.

```{r pip_vs_pvalue, fig.cap = "BQMA PIP versus limma p-value. Purple filled points are truly associated probes."}
plot_df <- data.frame(
  pip      = pip,
  neg_logp = -log10(limma_res[names(pip), "P.Value"]),
  is_true  = names(pip) %in% true_probes
)

plot(plot_df$neg_logp, plot_df$pip,
     col  = ifelse(plot_df$is_true, "purple", "grey60"),
     pch  = ifelse(plot_df$is_true, 19, 1),
     xlab = "-log10(p-value, limma)",
     ylab = "PIP (BQMA)",
     main = "BQMA PIP versus limma p-value",
     ylim = c(0, 1))
abline(h = 0.5,           lty = 2, col = "red",  lwd = 2)
abline(v = -log10(0.05),  lty = 2, col = "blue", lwd = 2)
legend("topleft",
       legend = c("Signal probe", "Null probe",
                  "PIP = 0.5", "p = 0.05"),
       col    = c("purple", "grey60", "red", "blue"),
       pch    = c(19, 1, NA, NA),
       lty    = c(NA, NA, 2, 2),
       bty    = "n")
```

## Annotating probes with QC flags

A high PIP is necessary but not sufficient for a reliable hit. Illumina
450k and EPIC arrays carry probes with known reliability issues:

- **SNPs at the CpG site (distance 0–2 bp)** — apparent methylation
  differences may reflect genotype rather than epigenetic state.
- **SNPs in the probe body (distance > 2 bp)** — can perturb
  hybridisation more subtly.
- **Cross-reactive probes** (Chen 2013 for 450k, McCartney 2016 for
  EPIC) — hybridise to multiple genomic locations and cannot be
  uniquely attributed.
- **Probes on chrX and chrY** — confound with sex unless explicitly
  modelled.

Because we used real probe identifiers from the 450k manifest in our
simulation, `annotate_results()` can look up each probe in the
annotation database and return its QC status.

`annotate_results()` returns a tidy data.frame combining the posterior
summaries with QC flags and genomic context (gene, feature, chr, pos).
Each probe is classified hierarchically:

- `CLEAN_CpG` — interpretable as epigenetic signal
- `FLAGGED_snp_body` — SNP in probe body (dist > 2 bp)
- `PROBLEMATIC` — SNP at CpG (dist ≤ 2 bp), cross-reactive, or on
  chrX / chrY

The individual underlying flags (`snp_dist_bp`, `cr_flag`, `xy_flag`)
are also returned so you can apply your own rule.

```{r annotate}
ann_df <- annotate_results(fit, array = "450k")

# One row per probe (single quantile for binary family)
nrow(ann_df)
head(ann_df[, c("probe", "pip", "beta_mean",
                "probe_class", "snp_dist_bp", "cr_flag", "xy_flag",
                "gene", "chr")])

# Class breakdown across all probes tested
table(ann_df$probe_class)
```

The exact mix depends on which 200 most-variable probes ended up in our
subset, but you should see a non-trivial number of `PROBLEMATIC` and
`FLAGGED_snp_body` probes alongside the `CLEAN_CpG` majority.

### Filtering top hits to reliable probes

The recommended workflow is: rank by PIP, then keep only probes
classified as `CLEAN_CpG` (or accept `FLAGGED_snp_body` if you want
broader coverage and report the flag).

```{r filter_clean}
ann_top <- ann_df[order(-ann_df$pip), ]
ann_top <- ann_top[ann_top$pip >= 0.5, ]

cat("Probes with PIP >= 0.5 :", nrow(ann_top), "\n")
cat("  CLEAN_CpG          :", sum(ann_top$probe_class == "CLEAN_CpG"),        "\n")
cat("  FLAGGED_snp_body   :", sum(ann_top$probe_class == "FLAGGED_snp_body"), "\n")
cat("  PROBLEMATIC        :", sum(ann_top$probe_class == "PROBLEMATIC"),      "\n")

# Reliable hits only
reliable <- ann_top[ann_top$probe_class == "CLEAN_CpG",
                    c("probe", "pip", "beta_mean", "ci_low", "ci_high",
                      "gene", "feature", "chr", "pos")]
print(reliable)
```

### Attaching the classification onto the fit

If you want the classification to travel with the `BQMAResult` object
(so that `show(fit)` prints the QC summary, and downstream functions
can read `@probe_class`), use `add_annotation()`. It is a thin writer
that runs `annotate_results()` internally and copies the per-probe
classification onto the slot.

```{r add_annotation}
fit <- add_annotation(fit, array = "450k")
show(fit)
```

The bottom line of `show(fit)` now reports the QC breakdown.

## Interpreting the results

The key outputs for downstream analysis are:

- `pip` values above 0.5 indicate probable association. Use 0.8 for
  a more stringent threshold analogous to genome-wide significance.
- `beta` values give the direction and magnitude of the association
  on the standardised M-value scale.
- `ci` values give the 95 % credible interval. A CI excluding zero is
  consistent with a non-null effect, but PIP is the primary evidence
  measure.
- `probe_class` from `annotate_results()` (or `@probe_class` after
  `add_annotation()`) flags reliability. Final reportable hits should
  generally restrict to `CLEAN_CpG`.

```{r final_summary}
final <- ann_df[ann_df$pip >= 0.5 &
                ann_df$probe_class == "CLEAN_CpG", ]
final <- final[order(-final$pip), ]

final$direction <- ifelse(final$beta_mean > 0,
                          "hypermethylated", "hypomethylated")

print(final[, c("probe", "pip", "beta_mean", "ci_low", "ci_high",
                "gene", "feature", "chr", "direction")])
```





## References

- Albert JH, Chib S (1993). Bayesian analysis of binary and
  polychotomous response data. *JASA* 88(422), 669–679.
- George EI, McCulloch RE (1993). Variable selection via Gibbs
  sampling. *JASA* 88(423), 881–889.
- Kozumi H, Kobayashi G (2011). Gibbs sampling methods for Bayesian
  quantile regression. *J. Stat. Comput. Simul.* 81(11), 1565–1578.
- Chen YA et al. (2013). Discovery of cross-reactive probes and
  polymorphic CpGs in the Illumina Infinium HumanMethylation450
  microarray. *Epigenetics* 8(2), 203–209.
- McCartney DL et al. (2016). Identification of polymorphic and
  off-target probe binding sites on the Illumina Infinium
  MethylationEPIC BeadChip. *Genomics Data* 9, 22–24.
- Hansen KD, Aryee M (2024). minfiData: Example data for the Illumina
  Methylation 450k array. R package.
